---
title: "BTRY 6020 Lab II"
output:
  pdf_document: default
  html_document: default
---

##Which Glucose measure to Use?

The following exercise (which uses a real data set) demonstrates all the points we have discussed in class about simple regression. The following points are covered:

1)	The usefulness of plotting both the data and the residuals to determine the fit of the regression model.
2)	The need to pay special attention to outliers and influential points.
3)	The use of transformations of the data to improve the fit and meet model assumptions.
4)	Use of tests and confidence intervals when the data have been transformed.

**Problem:** The concentration of glucose in the blood can be measured using a colorimeter and a substance that changes color in the presence of glucose. This test is highly sensitive, but is very vulnerable to contamination of the sample. Another test uses an enzyme, which is specific to glucose. This test is less vulnerable to contamination, has been used for years, and is highly accurate. The colorimetric test is less than one-tenth the cost of the other, and administrators wonder if they can switch over to the cheaper test. To examine the relationship between the results of these two tests, blood samples were taken from 46 patients at Stanford Hospital. The glucose concentration in each sample was measured using both methods. 

The main objective of the study is to determine how well the colorimetric values predict the enzymatic values. A prediction (regression) equation will be fitted, and the fit of the model examined.

A second objective is to determine a confidence interval for the mean enzymatic value of patients with colorimetric value 240, and a prediction interval for new patients with this colorimetric value.

The analysis begins by loading and plotting the data, and fitting the regression equation.

The data listed below are the two measures of glucose from blood samples taken from 46 patients at Stanford Medical Center in 1976.

```{r}
library(readxl)
ColEnz = read_excel("ColEnz.xlsx")
head(ColEnz)
```

The data are plotted below

```{r}
plot(ColEnz$Color, ColEnz$Enzyme, xlab = "Color", ylab = "Enzyme", main = "Color vs Enzyme relationship")
points(ColEnz$Color[16], ColEnz$Enzyme[16], lwd = 2, col = "red", pch = 7)
```

Based on the following linear model summary, answer the following questions.

```{r}
ColEnz.lm = lm(Enzyme~Color, data = ColEnz)
summary(ColEnz.lm)
```

a) What is the value of $R^2$? Does this indicate a good fit to the line?

b) Since both laboratory methods measure the same thing, the regression equation should have the form : $y = x + \epsilon$

    That is, $\beta_0 = 0$ and $\beta_1 =  1$. Do the data support this? (Do not worry about multiple comparisons here.)

```{r}
# hypothesis testing for intercep: HO: BO=0; HA: B0 not equal to zero 
tstat0 = (5.01364 - 0) /(3.47168)
tstat0
p_value0= 2*(1-pt(abs(tstat0), df=44))
p_value0
p_value0<0.05
```
We fail to reject H0, about that BO=0.

```{r}
# hypothesis testing for slope: HO: B1=1; HA: B1 not equal to one
tstat1 = (0.90905 - 1) /(0.02068)
tstat1
p_value1 = 2*(1-pt(abs(tstat1), df=44))
p_value1
p_value1<0.5
```

At 0.05 significance level, we reject the null hypothesis that the null=1.


c) The plot of residuals from this model against the predicted values is below. Does this plot have any statistically interesting features?

```{r}
plot(ColEnz.lm$fitted.values, ColEnz.lm$residuals)
points(ColEnz.lm$fitted.values[16], ColEnz.lm$residuals[16], lwd = 2, col = "red", pch = 7)
```

d)	The isolated point on this plot is patient 16. Patient 16 is also labeled on the scatterplot of the raw data. What should be done about this value?

The red point is a outlier, this patien was bad labelled.  

e) Actually, patient 16 was entered incorrectly into the computer. The colorimetric value of this patient should be 210, but was entered as 120. Below is the edited data set and the new linear regression fit. What effect did the error have on the regression equation? Why

```{r}
#save new data set
ColEnzU = ColEnz
#edit the colorimetric value of the 16th patient
ColEnzU[16,2] = 210

#fit the new regression
ColEnzU.lm = lm(Enzyme~Color, data = ColEnzU)
summary(ColEnzU.lm)
```

f)	What are the interesting features of the plot of residuals versus predicted values? What effect will this have on the estimate of the variance of $\beta_0$ vs. $\beta_1$? What effect will this have on the estimate of a confidence interval for the mean enzymatic value when the colorimetric value is 70? When it is 280?

```{r}
plot(ColEnzU.lm$fitted.values, ColEnzU.lm$residuals)
```

g) Suppose we consider the model $\log(y_i) = \beta_0 + \beta_1 \log(x_i) + \epsilon_i$.  What do we expect the values of $\beta_0$ and $\beta_1$ to be?

h)	The plot of residuals versus predicted values for this model are below. What are the interesting features of this plot?

```{r}
#log transformed linear regression
ColEnzT.lm = lm(log(Enzyme)~log(Color), data = ColEnzU)
#plotting  residuals vs predicted values
plot(ColEnzT.lm$fitted.values, ColEnzT.lm$residuals)
```

i)	The following plots shows the influence (Cook's distance) and leverage of our observations. What are the interesting features of these plots?

```{r} 
#Cook's Distance
plot(ColEnzT.lm, which = 4)

#Leverage Plot
lev = hat(model.matrix(ColEnzT.lm))
plot(lev, type="b", xlab="observation number", ylab="leverage", main="leverage plot")
```

j)	The isolated point on the residual plot is patient 11. Patient 11 has the highest leverage and Cook's distance. What is the effect of this data point on the regression equation? Checking against the patient chart shows that the correct data value was entered. Should we make any adjustment for this outlier?

k)	The investigators decided to fit the regression equation without patient 11. The following is the linear model summary.

```{r}
#remove 11th observations
ColEnzR = ColEnzU[-11,]

#fit linear model
ColEnzR.lm = lm(log(Enzyme)~log(Color), data = ColEnzR)
summary(ColEnzR.lm)
```

l)	The residual plot for this model now looks fine. Below are the leverage and influence plots. Are there any further problems of high leverage or influence for this model?

```{r}
#Cook's Distance
plot(ColEnzR.lm, which = 4)

#Leverage Plot
lev2 = hat(model.matrix(ColEnzR.lm))
plot(lev2, type="b", xlab="observation number", ylab="leverage", main="leverage plot")
```

m)	What is the estimate of the regression equation? What is a 95% confidence interval for $\beta_0$? For $\beta_1$?



n)	What is a 95% confidence interval for the mean enzymatic value when the colorimetric value is 240? ( $S_{xx}=13.41,  \bar{x}=4.83$) Please interpret.

o)	What is a 95% prediction interval for the enzymatic value of a patient whose colorimetric value is 240? Please interpret.









