---
title: "Lab 5 - Logistic Regression"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: cerulean
  word_document: default
---
-----

# Lab Goals

In this lab we will explore logistic regression models for binomial data in R.  In particular, we will examine:

  1. the `glm()` function

  2. interpreting coefficients
    
  3. diagnostics
    

# Cocaine Treatment - Binary Responses

An experiment was conducted to evaluate the effectiveness of two different drugs to prevent relapses in cocaine addiction.  A random sample of 72 former cocaine addicts were randomly assigned to one of three drug treatments (`Drug`).  One treatment group received the drug Lithium.  Another treatment group received the drug Desipramine.  The last treatment group received a placebo.  After a specified treatment period, the binary response `Relapse` was recorded where `Relapse = yes` indicates the individual returned to cocaine use and `Relapse = no` indicates the individual did not return to cocaine use.  The data can be found in the *CocaineTreatment.csv* file in the folder for Lab 5.   

Load this data set both into the console below and in this R Markdown document using the code chunk provided for you.  

```{r}
CocaineTreatment <- read.csv("CocaineTreatment.csv")
heads(CocaineTreatment )
```


## Data Organized as Bernoulli Trials

Here the data are considered as 72 bernoulli trials. We will run a logistic model with response `Relapse` and predictor `Drug`.  By default, the model estimates $p_i = P(Relapse_i = yes)$ (Why? It chooses the last level alphabetically as the "success" level).  However, you can change the ordering of the levels to estimate $p_i = P(Relapse_i = no)$.  Since we are interested in how well the drugs prevented relapse, we will change the ordering here.  Furthermore, we would like to have the placebo as the reference level for `Drug`.  So, we will also re-order `Drug` to make `Placebo` the first level. 

```{r}
# Change no to the last level of Relapse
CocaineTreatment$Relapse = factor(CocaineTreatment$Relapse, c("yes","no"))
#Change Placebo to the first level of Drug
CocaineTreatment$Drug = factor(CocaineTreatment$Drug, c("Placebo","Desipramine","Lithium"))
```


The `glm()` function will fit generalized linear models in R.  The logistic regression models are just one type of generalized linear models.  Here is the generic format of the `glm()` function:

--

glm(*formula*, family = *family*, data=*data*)

--

The arguments in *italics* need to be replaced by the specific information for your model. Where

  1) *formula* is of the form `Response ~ predictor1 + predictor2 + predictor3 + ...`

  2) for logistic regression *family* is `binomial`
  
  3) *data* is the name of your data set
  
Note: These are not the only arguments that are valid for this function.  We will explore this function more later.  
  
Here we will fit the model `coc1.glm` and run a summary of the output.

```{r}

#Fit the model
coc1.glm=glm(Relapse~Drug,family=binomial,data=CocaineTreatment)

#Summary
summary(coc1.glm)

```

1. Suppose this is an appropriate model for these data.  Does there appear to be any outliers? *None of the deviance residuals are greater than 2 in absolute value.  There doesn't appear to be any outliers.*

2. What is the fitted model for the treatment group that received Lithium?
  
  $log(\frac{\hat p_i}{1-\hat p_i}) = -1.61 + 0.511 = -1.099$
  
3. What is the fitted model for the treatment group that received Desipramine?
  
  $log(\frac{\hat p_i}{1-\hat p_i}) = -1.61 + 1.946 = .336$
  
4. For which drug was the estimated rate of relapse less than 50%? *Since 0.336 > 0, it is estimated that you have more than a 50% chance of not having a relapse if you take Desipramine.*
  
5. How can we interpret the partial slope associated with `DrugDesipramine`? *It is the estimated increase in the log odds of not having a relapse for a former addict that takes Desipamine in comparison to a former drug addict that takes a placebo.*

6. Using the summary information, perform a test to determine whether `Drug` is a significant predictor for this model.

  i) State the null and alternative hypotheses.
  
    *$H_0$: The partial slopes associated with Desipramine and Lithium are equal to 0*
    *$H_a$: At least one of the partial slopes associated with Desipramine and Lithium is not equal to 0*
    
  ii) The test statistic.

*To test whether the effects due to drug are significant, we should perform a likelihood ratio test (drop in deviance test) that compares a full model with `Drug` and an intercept only model.  The deviance for the intercept only model is labeled as the `Null deviance`.  The deviance for the full model is labeled as the `Residual Deviance`.  The test statistic is (`Null Deviance`-`Residual Deviance`).* 

```{r}
91.658-81.220
```

  iii) The reference distribution  *chi-square distribution with 2 degrees of freedom since 2 parameters were dropped from the model*
  
  iv) The following code will calculate the p-value for this test.  What do we conclude? *At the 0.01 significance level we conclude that `Drug` is a significant predictor for this model.*

```{r}
1-pchisq(10.438,2)
```

7. The `anova()` function will also perform this test.  First we will need to run the intercept only model.  The reduced (intercept only) model should be the first argument of `anova()`.  The second argument should be the full model.  You also need to include the option `test = "LRT"`.  Verify the p-value for this test is the same as the one calculated above.

```{r}
# Intercept only
cocint.glm=glm(Relapse~1,family=binomial,data=CocaineTreatment)
  
#LRT
anova(cocint.glm,coc1.glm,test="LRT")

```

## Data Organized as Binomial Observations

Since we have 24 independent observations for each drug treatment, we can also run the logistic model in R with Binomial observations.  The data set *CocaineTreatment2.csv* contains a Binomial observation for each level of `Treatment`.  The variable, `Yes`, is the number who relapsed on the drug treatment. `No` is the number who did not relapse on the drug treatment.  The total number of observations (trials) for each drug treatment is `Yes` + `No`.  

```{r}
CocaineTreatment2 <- read.csv("CocaineTreatment2.csv")
```


We can still use the `glm()` function to run a logistic regression model when the data follow a Binomial distribution.  In comparison to the run for Bernoulli data, the only change we need to make is to the *formula*.            

The format of *formula* should now be: `cbind(Successes,Failures) ~ predictor1 + predictor2 + ...` where
   
  i) `Successes` is the number of successes for each Binomial observation
  
  ii) `Failures` is the number of failures for each Binomial observation


Here we will refit the logistic regression model for predicting the probability of not having a relapse.  Again, the first step is to reorder the levels of `Treatment`.  Verify that the parameter estimates and Wald tests for this model match those from our first model.

```{r}
CocaineTreatment2$Treatment=factor(CocaineTreatment2$Treatment,c("Placebo","Desipramine","Lithium"))

coc2.glm=glm(cbind(No,Yes)~Treatment,family='binomial',data=CocaineTreatment2)

summary(coc2.glm)
```

Another way...

```{r}
coc3.glm=glm(No/(No+Yes)~Treatment,family=binomial,weights=(No+Yes),data=CocaineTreatment2)

summary(coc3.glm)
```



1. Why are the deviance residuals all 0? *Since we have three parameters for this model and three observations, the model fits each observation perfectly.*

2. What do you notice about the `Null deviance`?  *It is the same as the likelihood ratio test statistic for testing whether the choice of drug treatment significantly affects the probability of not having a relapse from our original model.* Has the LRT changed? *No.  The test statistic is the same and the number of degrees of freedom associated with the chi-square reference distribution has not changed.*

